---
title: "Handling Norwegian Characters (æ, ø, å) in R"
description: |
  This post explains how you can set the locale in R to handle Norwegian characters (æ, ø, å) properly.
author:
  - name: Øyvind Bugge Solheim
    url: https://www.oyvindsolheim.com
    affiliation: Institutt for samfunnsforskning (ISF)
    affiliation_url: https://www.samfunnsforskning.no
    orcid_id: 0000-0002-0871-5233
output: distill::distill_article
categories: 
  - locale
  - encoding
  - internationalization
  - R programming
date: "2024-12-20"
---


## **Introduction**  

This post was all wrong earlier. So I have rewritten it myself. Since "Bokmål" includes an "å", the locale was reverted to "C" by Rstudio. To make it work on a computer that has Norwegian Bokmål as the system language you have to specify the type of UTF-8 without the word "bokmål".^[I learned this here: https://github.com/rstudio/rstudio/issues/15008]. Thus the correct specification to work if you write in Norwegian is:

```{r}

Sys.setlocale(category = "LC_ALL", locale = "NO_NB.utf8")

```

For more information in Norwegian see here on my [home page](https://oyvindsolheim.com/code/norske%20bokstaver/).

# Put it in the startup of R

To make it standard you can put it in the r-profile. You open the profile:

```{r}

#install.packages("usethis")

usethis::edit_r_profile()

```

And enter the locale:

```{r}

Sys.setlocale(category = "LC_ALL", locale = "NO_NB.utf8")

```



